Distributional Trees and Forests

### 2017-11-22 pkg/disttree/inst/draft/draft.Rnw
draft without simulation section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1753 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6237ac5a31c1022e265155a69a9b7f8f546b3c9a

%% packages
\usepackage{amstext}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{thumbpdf}
\usepackage{rotating}
\usepackage{caption}
\captionsetup{format=hang}
\usepackage{subcaption}
\usepackage{Sweave}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{array, makecell}
\usepackage{tikz}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
%\usepackage[pagebackref=true]{hyperref}
%\usepackage{doi}

@@

\begin{figure}[t!]
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GLM, echo=FALSE, results=hide>>=
nobs <- 200
## GLM
set.seed(7)
x <- c(1:nobs)/nobs
ytrue <- 1+x*1.5
y <- ytrue + rnorm(nobs,0,0.3)
@

<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
\end{center}
\endminipage
{\LARGE$\rightarrow$}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAM, echo=FALSE, results=hide>>=
## GAM
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3 
y <- ytrue + rnorm(nobs,0,0.3)
@

<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
\end{center}
\endminipage
{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAMLSS, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(2*x)^2)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@

<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
\end{center}
\endminipage
\vspace{0.0cm}
\minipage{0.25\textwidth}
\begin{center}
\hspace{0.4cm}
LM, GLM 
\end{center}
\endminipage
\hspace{1.7cm}
\minipage{0.25\textwidth}
\begin{center}
GAM
\end{center}
\endminipage
\hspace{1.7cm}
\minipage{0.25\textwidth}
\begin{center}
GAMLSS
\end{center}
\endminipage
\caption{\label{Devel_parmod}Development in the field of parametric modeling.}
\end{figure}

\begin{figure}[t!]
\minipage{0.29\textwidth}
\begin{center}
<<motivation_regtree, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- yforest <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
y <- ytrue + rnorm(nobs,0,0.3)
for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
@

<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = ytree, col = pal["forest"], lwd=7)
@
\end{center}
\endminipage
{\LARGE$\rightarrow$}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_randforest, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
@

<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
@
\end{center}
\endminipage
{\LARGE$\rightarrow$}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_distforest, echo=FALSE, results=hide>>=
## distforest
set.seed(4723)
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
xt <- 2*(x-0.5)
var <- exp(-(2*xt)^4)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
x <- c(1:nobs)/nobs
@

<<plot_motivation_distforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
points(x, y, col = "slategray", pch = 19)
box(lwd=5)
@
\end{center}
\endminipage

\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
Regression Tree 
\end{center}
\endminipage
\hspace{0.65cm}
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
Random Forest 
\end{center}
\endminipage
\hspace{0.65cm}
\minipage{0.28\textwidth}
\begin{center}
\vspace{0.0cm}
Distributional Forest 
\end{center}
\endminipage\\
\minipage{0.29\textwidth}
\vspace{0.2cm}
\center {\LARGE$\downarrow$}\\
\vspace{-0.4cm}
%\hspace{0.0cm}
\begin{center}
<<motivation_disttree, echo=FALSE, results=hide>>=
## disttree
set.seed(74)
x <- c(1:nobs)/nobs
var <- ytree/3
y <- ytrue + rnorm(nobs,0,var)
@

<<plot_motivation_disttree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
lines(x = x, y = ytree, col = pal["forest"], lwd=7, main = "")
polygon(c(x, rev(x)), c(ytree + var, rev(ytree - var)),
  col = pallight["forest"], border = "transparent")
box(lwd=5)
@
\vspace{-0.2cm}\\
Distributional Tree
\vspace{0.2cm}
\end{center}
\endminipage
\caption{\label{Devel_treeforest}Development in the field of algorithmic trees and forests}
\end{figure}

@@

After explaining the novel methodology in Section~\ref{Methodology} of this paper and presenting results from a simulation study in Section~\ref{Simulation} the focus will then be put on an application in the field of probabilistic forecasting in Section~\ref{ProbabilisticForecasting}. The new method is applied on precipitation data in a mountainous region with a large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area. In a first setting only one observation station is taken into consideration. The models are trained on 24 years and prbabilistic predictions of the consecutive 4 years are validated. A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}. For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated by a forest model are plotted together with the corresponding observations. This illustrates that well fitting predicted expected values are provided by this model as they are close to the observed values.
Beside the expected values the predictions provided by distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.

@@

In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distributed with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left censored with censoring point 0. Therefore, the corresponding log-likelihood function with parameter vector $\theta = (\mu, \sigma)$ is

@@

\subsection{Distributional Tree}
\label{disttree}
Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$ the idea is to use the information provided by this model together with a set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ (from now on denoted by $\bold{Z}$ instead of $\bold{X}$ to indicate that they are considered as split variables only and not as regressors) to decide whether the data set should be split into subgroups or not. In particular, the information provided by the fitted model are the scores 
$$
s(\hat{\theta}, y_i) = \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i).
$$ 
These values express how well the model fits the data regarding each parameter. Ideally they should fluctuate randomly around zero, similar to residuals in ordinary least squares estimations, but in this case there is a score value for each pair of estimated parameter $\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$.\\
Statistical tests are applied to check whether there is a significant connection/dependency between the scores and each of the partitioning variables $Z_l \in \bold{Z} = (Z_1, \ldots, Z_m)$. If so, the data set is split into subgroups based on the values of the partitioning variable that shows the highest connection/dependency.\\
In a next step a local distributional model is fit in each of the subgroups. Repeating this procedure of fitting a model, applying statistical tests to evaluate the goodness of fit and splitting the data set depending on the test results leads to a tree structure. In that way the learning data set is separated into \textit{B} disjoint segments such that
$$
\mathcal{Z} = \dot{\bigcup\limits_{b = 1, \ldots, \textit{B}}} \mathcal{B}_b
$$
where each segment is represented by a terminal node of the tree.\\

@@

While the clear structure of such a tree model is advantageous in interpreting and analyzing it, the abrupt changes are often too rough and impose steps on the model even if the true underlying effect is smooth. In these situations bagging can solve the problem by combining an ensemble of trees such that wrongly detected abrupt changes of the model are turned into almost smooth transitions. In that way, all kinds of effects can be modeled, which is is also one the of the main advantages of forest models (taking bagging one step further as the variance is reduced by sampling the covariates considered for the splitting process in each node).

\subsection{Distributional Forest}
\label{distforest}
A distributional forest consists of an ensemble of distributional trees $\{t_s\}_{s=1,\ldots,T}$. Each of these trees is built on a different data set which can either be a bootstrap or a subset of the original data set. Moreover, in each node a subset of the partitioning variables is chosen randomly as candidates for splitting variables. In that way the correlation among the resulting trees is reduced and therefore also the variance of the model (\citealt{breiman2001random}).\\
\\
Forest models differ mainly in the way how the trees are combined to make predictions. In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process.\\
Contrary to distributional trees, in distributional forests the whole learning data set can be used to fit the distributional model and specify the predicted distribution for a new set of covariates. This difference is represented in the weight vector as the tree weights $\bold{\omega}^{tree}$ containing only zeros and ones are now replaced by integer valued forest weights \\
$\bold{\omega}^{forest} = (\omega^{forest}_1, \ldots, \omega^{forest}_n)$.\\
For a new set of covariates $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ the set of weights is obtained by the following procedure: For each observation $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}^{\ast}$. 
$$
\omega^{forest}_i(\bold{z}^{\ast}) = \sum_{s=1}^T \sum_{b=1}^{B^s} I((\bold{z}_i \in \mathcal{B}^s_b) \land (\bold{z}^{\ast} \in \mathcal{B}^s_b))
$$
where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.\\
This individual set of weights $\bold{\omega}^{forest}(\bold{z}^{\ast}) = (\omega^{forest}_1(\bold{z}^{\ast}), \ldots, \omega^{forest}_n(\bold{z}^{\ast}))$ for the new set of covariates $\bold{z}^{\ast}$ can now be included in the estimation process of the distribution parameters and in that way, leads to its individual parameter vector $\hat{\theta}(\bold{z}^{\ast})$.
$$
\hat{\theta}(\bold{z}^{\ast}) =  \max_{\theta \in \Theta} \sum_{i=1}^n \omega^{forest}_i(\bold{z}^{\ast}) \cdot \ell(\theta; y_i)
$$
While in a distributional tree the same distribution is specified for all sets of covariates being in the same node, a distributional forest predicts and fits a different distribution for each different set of covariates which is due to the individual weights calculated separately for each set of covariates.

\newpage
\section{Simulation Study}
\label{Simulation}
Within this simulation study the performance of distributional trees and forests is tested and compared to other models, showing situations where one or the other method is more advantageous than others.

@@

The new methods are evaluated on generated data sets and their results are then compared to those of a GAMLSS (\citealt{stasinopoulos2005}) which fits a separate generalized additive model to each of the distribution parameter and in that way specifies a whole distribution and is a very efficient option for smooth effects of the covariates on the parameters.\\
Each generated data set includes 500 samples consisiting of the value of a response variable $Y$ and six covariates $X_1, \ldots, X_6$ which are all independent and distributed as follows:
$$
X_1, X_2, X_3 \sim \textit{Unif}([-1,1])
$$
$$
X_4, X_5, X_6 \sim \textit{Binom}(1, 0.5)
$$

>>>>

%% packages
\usepackage{amstext}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{thumbpdf}
\usepackage{rotating}
\usepackage{caption}
\captionsetup{format=hang}
\usepackage{subcaption}
\usepackage{Sweave}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{array, makecell}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
%\usepackage[pagebackref=true]{hyperref}
%\usepackage{doi}

@@

\begin{figure}[t!]
\begin{tikzpicture}
\node (a) at (0,0){
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GLM, echo=FALSE, results=hide>>=
nobs <- 200
## GLM
set.seed(7)
x <- c(1:nobs)/nobs
ytrue <- 1+x*1.5
y <- ytrue + rnorm(nobs,0,0.3)
@

<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
\end{center}
\endminipage};
\node (b) at (5,0) 
{
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAM, echo=FALSE, results=hide>>=
## GAM
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3 
y <- ytrue + rnorm(nobs,0,0.3)
@

<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
\end{center}
\endminipage};
\node (c) at (10,0) 
{
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAMLSS, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(2*x)^2)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@

<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
\end{center}
\endminipage};
\node (d) at (0,-2) 
{
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
LM, GLM  
\end{center}
\endminipage};
\node (e) at (5,-2) 
{
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
GAM 
\end{center}
\endminipage};
\node at (10,-2)
{
\minipage{0.28\textwidth}
\begin{center}
\vspace{0.0cm}
GAMLSS 
\end{center}
\endminipage};
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
\end{tikzpicture}
\caption{\label{Devel_parmod}Development in the field of parametric modeling.}
\end{figure}

\begin{figure}[t!]
\begin{tikzpicture}
\node (a) at (0,0){
\minipage{0.29\textwidth}
\begin{center}
<<motivation_regtree, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- yforest <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
y <- ytrue + rnorm(nobs,0,0.3)
for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
@

<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = ytree, col = pal["forest"], lwd=7)
@
\end{center}
\endminipage};
\node (b) at (5,0) 
{
\minipage{0.29\textwidth}
\begin{center}
<<motivation_randforest, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
@

<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
@
\end{center}
\endminipage};
\node (c) at (10,0) 
{
\minipage{0.29\textwidth}
\begin{center}
<<motivation_distforest, echo=FALSE, results=hide>>=
## distforest
set.seed(4723)
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
xt <- 2*(x-0.5)
var <- exp(-(2*xt)^4)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
x <- c(1:nobs)/nobs
@

<<plot_motivation_distforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
points(x, y, col = "slategray", pch = 19)
box(lwd=5)
@
\end{center}
\endminipage};
\node (d) at (0,-2) 
{
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
Regression Tree 
\end{center}
\endminipage};
\node (e) at (5,-2) 
{
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
Random Forest 
\end{center}
\endminipage};
\node at (10,-2)
{
\minipage{0.28\textwidth}
\begin{center}
\vspace{0.0cm}
Distributional Forest 
\end{center}
\endminipage};
\node (g) at (0,-5.4) 
{
\minipage{0.29\textwidth}
\vspace{0.2cm}
%\hspace{0.0cm}
\begin{center}
<<motivation_disttree, echo=FALSE, results=hide>>=
## disttree
set.seed(74)
x <- c(1:nobs)/nobs
var <- ytree/3
y <- ytrue + rnorm(nobs,0,var)
@

<<plot_motivation_disttree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
lines(x = x, y = ytree, col = pal["forest"], lwd=7, main = "")
polygon(c(x, rev(x)), c(ytree + var, rev(ytree - var)),
  col = pallight["forest"], border = "transparent")
box(lwd=5)
@
\vspace{-0.2cm}%\\
Distributional Tree
\vspace{0.2cm}
\end{center}
\endminipage};
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](0,-2.4)--(0,-3.2);
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.2,-5.4)--(7.9,-2.5);
\end{tikzpicture}
\caption{\label{Devel_treeforest}Development in the field of algorithmic trees and forests}
\end{figure}

%\SweaveOpts{eval = FALSE}

@@

After explaining the novel methodology in Section~\ref{Methodology} of this paper the focus will then be put on an application in the field of probabilistic forecasting in Section~\ref{ProbabilisticForecasting}. The new method is applied on precipitation data in a mountainous region with a large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area. In a first setting only one observation station is taken into consideration. The models are trained on data from 24 years and probabilistic predictions of the consecutive 4 years are validated. A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}. For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated by a forest model are plotted together with the corresponding observations. This illustrates that well fitting predicted expected values are provided by this model as they are close to the observed values.
Beside the expected values the predictions provided by distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis.

@@

In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distributed with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left censored with censoring point 0. (The power transformation is used to remove large parts of the skewness of the data as further explained in \ref{setting}.) Therefore, the corresponding log-likelihood function with parameter vector $\theta = (\mu, \sigma)$ is

@@

\subsection{Distributional Tree}
\label{disttree}
Given the fit of a global distributional model $D(Y, \bold{\hat{\theta}})$ the idea is to use the information provided by this model together with a set of covariates $\bold{Z} = (Z_1, \ldots, Z_m) \in \mathcal{Z}$ (from now on denoted by $\bold{Z}$ instead of $\bold{X}$ to indicate that they are considered as split variables only and not as regressors) to decide whether the data set should be split into subgroups or not. In particular, the information provided by the fitted model are the scores 
$$
s(\hat{\theta}, y_i) = \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i).
$$ 
These values express how well the model fits the data regarding each parameter. Ideally they should fluctuate randomly around zero, similar to residuals in ordinary least squares estimations, but in this case there is a score value for each pair of estimated parameter $\hat{\theta}_j \in (\hat{\theta}_1, \ldots, \hat{\theta}_k) = \bold{\hat{\theta}}$ and observation $y_j \in \{y_i\}_{i = 1,\ldots, n}$. This enables the model to be sensitive to changes in each of the parameters which is one of the main advantages of distributional trees over other tree models considering only the mean of the distribution.\\
Statistical tests are applied to check whether there is a significant connection/dependency between the scores and each of the partitioning variables $Z_l \in \bold{Z} = (Z_1, \ldots, Z_m)$. If so, the data set is split into subgroups based on the values of the partitioning variable that shows the highest connection/dependency.\\
In a next step a local distributional model is fit in each of the subgroups. Repeating this procedure of fitting a model, applying statistical tests to evaluate the goodness of fit and splitting the data set depending on the test results leads to a tree structure. In that way the learning data set is separated into \textit{B} disjoint segments such that
$$
\mathcal{Z} = \dot{\bigcup\limits_{b = 1, \ldots, \textit{B}}} \mathcal{B}_b
$$
where each segment is represented by a terminal node of the tree.\\

@@

While the clear structure of such a tree model is advantageous in interpreting and analyzing it, the abrupt changes are often too rough and impose steps on the model even if the true underlying effect is smooth. In these situations bagging can solve the problem by combining an ensemble of trees such that wrongly detected abrupt changes of the model are turned into almost smooth transitions. In that way, all kinds of effects can be modeled, which is is also one the of the main advantages of forest models (taking bagging one step further as the variance is reduced by sampling the covariates considered for the splitting process in each node). \\
\\
Common forest models such as random forests consist of trees only considering the mean of the distribution of the response variable. For that reason, changes in other parameters can only be detected if they are connected to changes in the location parameter, e.g. if they depend on the location parameter. Therefore, building a forest of distributional trees which can also capture separate changes in each of the distribution parameters leads to an improvement of the model. This valuable feature of detecting separate changes is also provided by GAMLSS as each parameter is modeled separately. However, in this case this separation can also be a disadvantage as it hinders the capturing of interactions unless they are specified explicitly by the user. These two examples/situations provide once again a good reason to combine the two frameworks of flexible parametric modeling and tree structured models in order to profit from their advantages and compensate possible disadvantages. This is the pursued goal in introducing distributional forests.

\subsection{Distributional Forest}
\label{distforest}
A distributional forest consists of an ensemble of distributional trees $\{t_s\}_{s=1,\ldots,T}$. Each of these trees is built on a different data set which can either be a bootstrap or a subset of the original data set. Moreover, in each node a subset of the partitioning variables is chosen randomly as candidates for splitting variables. In that way the correlation among the resulting trees is reduced and therefore also the variance of the model (\citealt{breiman2001random}).\\
\\
Forest models differ mainly in the way how the trees are combined to make predictions. In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process.\\
Contrary to distributional trees, in distributional forests the whole learning data set can be used to fit the distributional model and specify the predicted distribution for a new set of covariates. This difference is represented in the weight vector as the tree weights $\bold{\omega}^{tree}$ containing only zeros and ones are now replaced by integer valued forest weights \\
$\bold{\omega}^{forest} = (\omega^{forest}_1, \ldots, \omega^{forest}_n)$.\\
For a new set of covariates $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ the set of weights is obtained by the following procedure: For each observation $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}^{\ast}$. 
$$
\omega^{forest}_i(\bold{z}^{\ast}) = \sum_{s=1}^T \sum_{b=1}^{B^s} I((\bold{z}_i \in \mathcal{B}^s_b) \land (\bold{z}^{\ast} \in \mathcal{B}^s_b))
$$
where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.\\
This individual set of weights $\bold{\omega}^{forest}(\bold{z}^{\ast}) = (\omega^{forest}_1(\bold{z}^{\ast}), \ldots, \omega^{forest}_n(\bold{z}^{\ast}))$ for the new set of covariates $\bold{z}^{\ast}$ can now be included in the estimation process of the distribution parameters and in that way, leads to its individual parameter vector $\hat{\theta}(\bold{z}^{\ast})$.
$$
\hat{\theta}(\bold{z}^{\ast}) =  \max_{\theta \in \Theta} \sum_{i=1}^n \omega^{forest}_i(\bold{z}^{\ast}) \cdot \ell(\theta; y_i)
$$
While in a distributional tree the same distribution is specified for all sets of covariates being in the same node, a distributional forest predicts and fits a different distribution for each different set of covariates which is due to the individual weights calculated separately for each set of covariates.\\
\\
As this novel methodology combines the advantages of smooth and flexible parametric modeling with those of tree structured modeling it provides a model that can deal with a wide range of situations. Abrupt changes and non-additive effects can be captured by its tree structure while smooth effects can be modeled reasonably by combining various tree and smoothing the steps imposed by a single tree. As scores are considered for each distribution parameter (separate) effects in different parameters can be captured. At the same time, also interactions are detected automatically by splitting the data in a way that is sensitive to changes in either each of the parameters or combined in various parameters.\\
Even though many modeling situations demand for the use of a specific model providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance what kind of situation this is. In these cases a distributional forest is a recommendable compromise providing reasonable models and performing well without any prespecification or additional knowledge needed in advance.

\SweaveOpts{eval = FALSE}
%\newpage
%\section{Simulation Study}
%\label{Simulation}
%Within this simulation study the performance of distributional trees and forests is tested and compared to other models, showing situations where one or the other method is more advantageous than others.

@@

%The new methods are evaluated on generated data sets and their results are then compared to those of a GAMLSS (\citealt{stasinopoulos2005}) which fits a separate generalized additive model to each of the distribution parameter and in that way specifies a whole distribution and is a very efficient option for smooth effects of the covariates on the parameters.\\
%Each generated data set includes 500 samples consisiting of the value of a response variable $Y$ and six covariates $X_1, \ldots, X_6$ which are all independent and distributed as follows:
%$$
%X_1, X_2, X_3 \sim \textit{Unif}([-1,1])
%$$
%$$
%X_4, X_5, X_6 \sim \textit{Binom}(1, 0.5)
%$$

%The distribution of the response variable is a left censored normal distribution with censoring point 0. The corresponding distribution parameters are defined by a parameter function $f$ depending on the covariates and an additional factor $\kappa$, also known as the smoothness constraint, which regulates the steepness of certain sections of the parameter function. For low values of $\kappa$ the resulting parameters change rather slowly over varying values of the covariates while high values of $\kappa$ lead to abrupt changes. \\ 
%The location parameter $\mu$ depends on $X_1, X_4$ and $\kappa$ and the scale parameter $\sigma$ depends on $\mu, X_1$ and $\kappa$. The other covariates $X_2, X_3, X_5$ and $X_6$ are used as noise variables.\\
%The latent variable $Y^{\ast}$ is generated based on the following setting.
%\begin{center}
%$$ 
%Y^{\ast} \sim \mathcal{N} (f(X_1, X_4, \kappa))
%$$
%with
%$$
%f(X_1, X_4, \kappa) = \left(\mu(X_1, X_4, \kappa) , \sigma(\mu,X_1,\kappa)\right)
%$$ 
%and
%\begin{align*}
%\mu(X_1, X_4, \kappa) = 0.2 
%+ \begin{cases}
%    7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})},
%    & \text{if } X_1 < 0.4 \text{ and } X_4 = 0\\
%    7 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{7}{2},       
%    & \text{if } X_1 \geq 0.4 \text{ and } X_4 = 0\\
%    10 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})},
%    & \text{if } X_1 < 0.4 \text{ and } X_4 = 1\\
%    13 \cdot \exp{(-(3 \cdot X_1-1.2)^{(2\cdot \kappa)})} \cdot \frac{1}{2} + \frac{7}{2},       
%    & \text{if } X_1 \geq 0.4 \text{ and } X_4 = 1
%\end{cases}
%\end{align*}

%$$
%\sigma(\mu, X_1,\kappa) = 0.5 + \frac{\mu}{5} + 2 \cdot \mathbf{1}(X_1 < -0.4)
%$$
%%where $\Lambda$ is the cumulative distribution function of the logistic distribution
%%$$
%%\Lambda(\omega) = \frac{1}{1 + \exp(-\omega)}
%%$$
%The response variable $Y$ is defined as 
%$$
%Y = 
%\begin{cases}
%    Y^{\ast},& \text{if } Y^{\ast} \geq 0\\
%    0,       & \text{else}
%\end{cases}
%$$
%\end{center}
